# Risk of recidivism analysis

## Setup 

```{r}
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
 tidyverse, # tidyverse packages 
 conflicted, # an alternative conflict resolution strategy 
 patchwork, # arranging ggplots
 scales, # rescaling 
 survival, # survival analysis
 broom, # for modeling
 here, # reproducibility 
 glue # pasting strings and objects 
)

# To avoid conflicts 
conflict_prefer("filter", "dplyr") 
conflict_prefer("select", "dplyr") 
```

## Load data 

We select fields for severity of charge, number of priors, demographics, age, sex, COMPAS scores, and whether each person was accused of a crime within two years.

- N of observations (rows): 7,214
- N of variables (columns): 53

```{r message=FALSE}
two_years <- read_csv(here("data", "compas-scores-two-years.csv"))

dim(two_years)
```

## Wrangling 

- Not all of the observations are useable for the first round of analysis.
- There are a number of reasons to remove rows because of missing data:
    - If the charge date of a defendants COMPAS scored crime was not within 30 days from when the person was arrested, we assume that because of data quality reasons, that we do not have the right offense.
    - We coded the recidivist flag -- is_recid -- to be -1 if we could not find a COMPAS case at all.
    - In a similar vein, ordinary traffic offenses -- those with a c_charge_degree of 'O' -- will not result in Jail time are removed (only two of them).
    - We filtered the underlying data from Broward county to include only those rows representing people who had either recidivated in two years, or had at least two years outside of a correctional facility.

### Create a function 

```{r}

wrangle_data <- function(data){

df <- data %>% 
    
    # Select variables 
    select(age, c_charge_degree, race, age_cat, score_text, sex, priors_count, days_b_screening_arrest, decile_score, is_recid, two_year_recid, 
         c_jail_in, c_jail_out) %>% 
    # Filter rows 
    filter(days_b_screening_arrest <= 30,
           days_b_screening_arrest >= -30, 
           is_recid != -1,
           c_charge_degree != "O",
           score_text != 'N/A') %>% 
    # Mutate variables 
    mutate(length_of_stay = as.numeric(as.Date(c_jail_out) - as.Date(c_jail_in)),
           c_charge_degree = factor(c_charge_degree),
           age_cat = factor(age_cat),
           race = factor(race, levels = c("Caucasian","African-American","Hispanic","Other","Asian","Native American")),
           sex = factor(sex, levels = c("Male","Female")),
           score_text = factor(score_text, levels = c("Low", "Medium", "High")),
           score = score_text,
# I added this new variable to test whether measuring the DV as a binary or continuous var makes a difference 
           score_num = as.numeric(score_text)) %>% 
    # Rename variables 
    rename(crime = c_charge_degree,
           gender = sex)
        
return(df)}
```

### Apply the function to the data 

```{r}
df <- wrangle_data(two_years)

names(df)

# Check whether the function works as expected
head(df, 5)
```

## Descriptive analysis 

- Higher COMPAS scores are slightly correlated with a longer length of stay.

```{r}
cor(df$length_of_stay, df$decile_score)

# Set theme
ggplot2::theme_set(theme_minimal())

df %>%
  group_by(score) %>%
  count() %>%
  ggplot(aes(x = score, y = n)) +
    geom_col() +
    labs(x = "Score",
         y = "Count",
         title = "Score distribution")
```
Judges are often presented with two sets of scores from the COMPAS system -- one that classifies people into High, Medium and Low risk, and a corresponding decile score. There is a clear downward trend in the decile scores as those scores increase for white defendants.

```{r}
df %>%
  ggplot(aes(ordered(decile_score))) + 
          geom_bar() +
          facet_wrap(~race, nrow = 2) +
          labs(x = "Decile Score",
               y = "Count",
               Title = "Defendant's Decile Score")
```

## Modeling 

After filtering out bad rows, our first question is whether there is a significant difference in COMPAS scores between races. To do so we need to change some variables into factors, and run a logistic regression, comparing low scores to high scores.

### Model building 

```{r}

model_data <- function(data){

# Logistic regression model
lr_model <- glm(score ~ gender + age_cat + race + priors_count + crime + two_year_recid, 
             family = "binomial", data = data)

# OLS, DV = score_num
ols_model1 <- lm(score_num ~ gender + age_cat + race + priors_count + crime + two_year_recid, data = data)

# OLS, DV = decile_score 
ols_model2 <- lm(decile_score ~ gender + age_cat + race + priors_count + crime + two_year_recid, data = data)

# Extract model outcomes with confidence intervals 
lr_est <- lr_model %>% 
    tidy(conf.int = TRUE) 

ols_est1 <- ols_model1 %>%
    tidy(conf.int = TRUE) 

ols_est2 <- ols_model2 %>%
    tidy(conf.int = TRUE) 

# AIC scores 
lr_AIC <- AIC(lr_model)
ols_AIC1 <- AIC(ols_model1)
ols_AIC2 <- AIC(ols_model2)
    
list(lr_est, ols_est1, ols_est2, 
     lr_AIC, ols_AIC1, ols_AIC2)

}

```

### Model comparisons 

```{r}

glue("AIC score of logistic regression: {model_data(df)[4]} 
      AIC score of OLS regression (with categorical DV):  {model_data(df)[5]}
      AIC score of OLS regression (with continuous DV): {model_data(df)[6]}")

```

### Logistic regression model 

```{r}
lr_model <- model_data(df)[1] %>% data.frame()

lr_model %>%
  filter(term != "(Intercept)") %>%
  mutate(term = gsub("race|age_cat|gender|M","", term)) %>%
  ggplot(aes(x = fct_reorder(term, estimate), y = estimate, ymax = conf.high, ymin = conf.low)) +
  geom_pointrange() +
  coord_flip() +
  labs(y = "Estimate", x = "",
      title = "Logistic regression") +
  geom_hline(yintercept = 0, linetype = "dashed")

```

```{r}
interpret_estimate <- function(model){
    
    # Control 
    intercept <- model$estimate[model$term == "(Intercept)"]
    control <- exp(intercept) / (1 + exp(intercept))
    
    # Likelihood 
    model <- model %>% filter(term != "(Intercept)")
    
    model$likelihood <- (exp(model$estimate) / (1 - control + (control * exp(model$estimate))))
    
    return(model)
}
```

```{r}
interpret_estimate(lr_model) %>%
    mutate(term = gsub("race|age_cat|gender","", term)) %>% 
    ggplot(aes(x = fct_reorder(term, likelihood), y = likelihood)) +
        geom_point(size = 3) +
        coord_flip() +
        labs(y = "Likelihood", x = "",
            title ="Logistic regression") +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        geom_hline(yintercept = 1, linetype = "dashed")
```









