# Risk of Violent Recidivism

## Setup 

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
 tidyverse, # tidyverse packages 
 conflicted, # an alternative conflict resolution strategy 
 ggthemes, # other themes for ggplot2 
 patchwork, # arranging ggplots
 scales, # rescaling 
 survival, # survival analysis
 broom, # for modeling
 here, # reproducibility 
 glue # pasting strings and objects 
)

# To avoid conflicts 
conflict_prefer("filter", "dplyr") 
conflict_prefer("select", "dplyr") 

# Set themes 
theme_set(ggthemes::theme_fivethirtyeight())
```

## Load data 

```{r}
two_years_violent <- read_csv(here("data" ,"compas-scores-two-years-violent.csv"))

glue("N of observations (rows): {nrow(two_years_violent)}
      N of variables (columns): {ncol(two_years_violent)}")

```

## Wrangling

### Create a function 

```{r}

wrangle_data <- function(data){

df <- data %>% 
    
    # Select variables 
    select(age, c_charge_degree, race, age_cat, v_score_text, sex, priors_count, 
         days_b_screening_arrest, v_decile_score, is_recid, two_year_recid) %>%            
    # Filter rows 
    filter(days_b_screening_arrest <= 30,
           days_b_screening_arrest >= -30, 
           is_recid != -1,
           c_charge_degree != "O",
           v_score_text != 'N/A') %>% 
    # Mutate variables 
    mutate(c_charge_degree = factor(c_charge_degree),
           age_cat = factor(age_cat),
           race = factor(race, levels = c("Caucasian","African-American","Hispanic","Other","Asian","Native American")),
           sex = factor(sex, levels = c("Male","Female")),
           v_score_text = factor(v_score_text, levels = c("Low", "Medium", "High")),
# I added this new variable to test whether measuring the DV as a binary or continuous var makes a difference 
           score_num = as.numeric(v_score_text)) %>%
    # Rename variables 
    rename(crime = c_charge_degree,
           gender = sex,
           score = v_score_text)
        
return(df)}

```

### Apply the function to the data 

```{r}

df <- wrangle_data(two_years_violent)

names(df)

head(df, 5) # Check whether the function works as expected 

```

### Descriptive analysis 

Score distribution 

```{r}
df %>%
  group_by(score) %>%
  count() %>%
  ggplot(aes(x = score, y = n)) +
    geom_col() +
    labs(x = "Score",
         y = "Count",
         title = "Score distribution")
```

Score distribution by race

```{r}

df %>%
  ggplot(aes(ordered(v_decile_score))) + 
          geom_bar() +
          facet_wrap(~race, nrow = 2) +
          labs(x = "Decile Score",
               y = "Count",
               Title = "Defendant's Decile Score")

```

### Modeling 

After filtering out bad rows, our first question is whether there is a significant difference in COMPAS scores between races. To do so we need to change some variables into factors, and run a logistic regression, comparing low scores to high scores.

```{r}

model_data <- function(data){

# Logistic regression model
lr_model <- glm(score ~ gender + age_cat + race + priors_count + crime + two_year_recid, 
             family = "binomial", data = data)

# OLS
ols_model1 <- lm(score_num ~ gender + age_cat + race + priors_count + crime + two_year_recid, 
             data = data)

ols_model2 <- lm(v_decile_score ~ gender + age_cat + race + priors_count + crime + two_year_recid, 
             data = data)

# Extract model outcomes with confidence intervals 
lr_est <- lr_model %>% 
    tidy(conf.int = TRUE) 

ols_est1 <- ols_model1 %>%
    tidy(conf.int = TRUE) 

ols_est2 <- ols_model2 %>%
    tidy(conf.int = TRUE) 

# AIC scores 
lr_AIC <- AIC(lr_model)
ols_AIC1 <- AIC(ols_model1)
ols_AIC2 <- AIC(ols_model2)
    
list(lr_est, ols_est1, ols_est2, lr_AIC, ols_AIC1, ols_AIC2)
}
    
```

#### Model comparisons 

```{r}

glue("AIC score of logistic regression: {model_data(df)[4]} 
      AIC score of OLS regression (with categorical DV):  {model_data(df)[5]}
      AIC score of OLS regression (with continuous DV): {model_data(df)[6]}")

```

#### Logistic regression model 

```{r}

lr_model <- model_data(df)[1] %>% 
  data.frame()

lr_model %>%
  filter(term != "(Intercept)") %>%
  mutate(term = gsub("race|age_cat|gender","", term)) %>%
  ggplot(aes(x = fct_reorder(term, estimate), y = estimate, ymax = conf.high, ymin = conf.low)) +
  geom_pointrange() +
  coord_flip() +
  labs(y = "Estimate", x = "",
      title = "Logistic regression") +
  geom_hline(yintercept = 0, linetype = "dashed")

```

```{r}

interpret_estimate <- function(model){
    
    # Control 
    intercept <- model$estimate[model$term == "(Intercept)"]
    control <- exp(intercept) / (1 + exp(intercept))
    
    # Likelihood 
    model <- model %>% filter(term != "(Intercept)")
    
    model$likelihood <- (exp(model$estimate) / (1 - control + (control * exp(model$estimate))))
    
    return(model)
}

interpret_estimate(lr_model) %>%
    mutate(term = gsub("race|age_cat|gender","", term)) %>% 
    ggplot(aes(x = fct_reorder(term, likelihood), y = likelihood)) +
        geom_point(size = 3) +
        coord_flip() +
        labs(y = "Likelihood", x = "",
            title ="Logistic regression") +
        scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
        geom_hline(yintercept = 1, linetype = "dashed")

```









