# Bias in the algorithm 

- In order to test whether COMPAS scores do an accurate job of deciding whether an offender is Low, Medium or High risk, we ran a Cox Proportional Hazards model. Northpointe, the company that created COMPAS and markets it to Law Enforcement, also ran a Cox model in [their validation study](https://journals.sagepub.com/doi/abs/10.1177/0093854808326545).

- We used the counting model and removed people when they were incarcerated. Due to errors in the underlying jail data, we need to filter out 32 rows that have an end date more than the start date. Considering that there are 13,334 total rows in the data, such a small amount of errors will not affect the results.

## Setup 

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
 tidyverse, # tidyverse packages 
 conflicted, # an alternative conflict resolution strategy 
 ggthemes, # other themes for ggplot2 
 patchwork, # arranging ggplots
 scales, # rescaling 
 survival, # survival analysis
 broom, # for modeling
 here, # reproducibility 
 glue, # pasting strings and objects 
 reticulate # source python codes
)

# To avoid conflicts 
conflict_prefer("filter", "dplyr") 
conflict_prefer("select", "dplyr") 

# Set themes 
theme_set(ggthemes::theme_fivethirtyeight())

```


## Load data 

```{r}
cox_data <- read_csv(here("data" ,"cox-parsed.csv"))

glue("N of observations (rows): {nrow(cox_data)}
      N of variables (columns): {ncol(cox_data)}")

```

## Wrangling

```{r}

df <- cox_data %>% 
    filter(score_text != "N/A") %>%
    filter(end > start) %>%
    mutate(c_charge_degree = factor(c_charge_degree),
           age_cat = factor(age_cat),
           race = factor(race, levels = c("Caucasian","African-American","Hispanic","Other","Asian","Native American")),
           sex = factor(sex, levels = c("Male","Female")),
           score_factor = factor(score_text, levels = c("Low", "Medium", "High")))

grp <- df[!duplicated(df$id),]

```

### Descriptive analysis 

Score distribution 

```{r}
grp %>% 
    group_by(score_factor) %>%
      count() %>%
      ggplot(aes(x = score_factor, y = n)) +
        geom_col() +
        labs(x = "Score",
             y = "Count",
             title = "Score distribution")
```

Score distribution by race

```{r}

df %>%
  ggplot(aes(ordered(score_factor))) + 
          geom_bar() +
          facet_wrap(~race, nrow = 2) +
          labs(x = "Decile Score",
               y = "Count",
               Title = "Defendant's Decile Score")

```

## Modeling 

```{r}

f2 <- Surv(start, end, event, type="counting") ~ race + score_factor + race * score_factor

model <- coxph(f2, data = df)

model %>%
  broom::tidy(conf.int = TRUE) %>%
  mutate(term = gsub("race|score_factor","", term)) %>% 
  filter(term != "<chr>") %>%
  ggplot(aes(x = fct_reorder(term, estimate), y = estimate, ymax = conf.high, ymin = conf.low)) +
  geom_pointrange() +
  coord_flip() +
  labs(y = "Estimate", x = "")
    
```

The interaction term shows a similar disparity as the logistic regression above.

High risk white defendants are 3.61 more likely than low risk white defendants, while High risk black defendants are 2.99 more likely than low.

```{r}
visualize_surv <- function(input){
  
f <- Surv(start, end, event, type="counting") ~ score_factor

fit <- survfit(f, data = input)

fit %>%
    tidy(conf.int = TRUE) %>%
    mutate(strata = gsub("score_factor=","", strata)) %>%
    mutate(strata = factor(strata, levels = c("High","Medium","Low"))) %>%
    ggplot(aes(x = time, y = estimate, ymax = conf.high, ymin = conf.low, group = strata, col = strata)) +
    geom_pointrange(alpha = 0.1) +
    guides(colour = guide_legend(override.aes = list(alpha = 1))) +
    ylim(c(0, 1)) +
    labs(x = "Time", y = "Estimated survival rate", col = "Strata")}
```

```{r}
visualize_surv(df) + ggtitle("Overall")
```

Black defendants do recidivate at higher rates according to race specific Kaplan Meier plots.

```{r}

(df %>% filter(race == "Caucasian") %>% visualize_surv() + ggtitle("Caucasian")) /
(df %>% filter(race == "African-American") %>% visualize_surv() + ggtitle("African-American")) 

```

In terms of underlying recidivism rates, we can look at gender specific Kaplan Meier estimates. There is a striking difference between women and men.

```{r}

(df %>% filter(sex == "Female") %>% visualize_surv() + ggtitle("Female")) /

(df %>% filter(sex == "Male") %>% visualize_surv() + ggtitle("Male"))

```

As these plots show, the COMPAS score treats a High risk women the same as a Medium risk man.

### Risk of Recidivism accuracy 

The above analysis shows that the COMPAS algorithm does overpredict African-American defendant's future recidivism, but we haven't yet explored the direction of the bias. We can discover fine differences in overprediction and underprediction by comparing COMPAS scores across racial lines.

```{r cache=TRUE}

# create a new environment 
conda_create("r-reticulate")

# install libs 
conda_install("r-reticulate", c("pandas"))

# indicate that we want to use a specific condaenv
use_condaenv("r-reticulate")

```


```{python}

from truth_tables import PeekyReader, Person, table, is_race, count, vtable, hightable, vhightable
from csv import DictReader

people = []

```

```{python}

with open("./data/cox-parsed.csv") as f:
    reader = PeekyReader(DictReader(f))
    try:
        while True:
            p = Person(reader)
            if p.valid:
                people.append(p)
    except StopIteration:
        pass

```

```{python}

pop = list(filter(lambda i: ((i.recidivist == True and i.lifetime <= 730) or
                              i.lifetime > 730), list(filter(lambda x: x.score_valid, people))))

recid = list(filter(lambda i: i.recidivist == True and i.lifetime <= 730, pop))

rset = set(recid)

surv = [i for i in pop if i not in rset]

```

Define a function for a table.

```{python}

import pandas as pd 

def create_table(x, y):

  t = table(list(x), list(y))
  
  df = pd.DataFrame(t.items(), 
             columns = ['Metrics', 'Scores'])
             
  return(df)
             
```

- All defenders 

```{python}

create_table(list(recid), list(surv)).to_csv("data/table_recid.csv")

``` 

```{r}

read.csv(here("data", "table_recid.csv"))[,-1] %>%
  ggplot(aes(x = Metrics, y = Scores)) +
  geom_col() +
  labs(title = "Recidivism")

```

That number is higher for African Americans at 44.85% and lower for whites at 23.45%.

```{python}

def create_comp_tables(recid_data, surv_data):
  
    # filtering variables 
    is_afam = is_race("African-American")
    is_white = is_race("Caucasian")
  
    # dfs 
    df1 = create_table(filter(is_afam, recid_data),
                       filter(is_afam, surv_data))
  
    df2 = create_table(filter(is_white, recid_data), 
                       filter(is_white, surv_data))
  
    # concat 
    dfs = pd.concat([df1, df2])
    
    dfs['Group'] = ['African Americans','African Americans','Whites','Whites']
    
    return(dfs)
    
```

```{python}

create_comp_tables(recid, surv).to_csv("data/comp_tables_recid.csv")

```

```{r}

read.csv(here("data", "comp_tables_recid.csv"))[,-1] %>%
  ggplot(aes(x = Metrics, y = Scores, fill = Group)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Recidivism")

```

### Risk of Violent Recidivism accuracy

COMPAS also offers a score that aims to measure a persons risk of violent recidivism, which has a similar overall accuracy to the Recidivism score.

```{python}

vpeople = []

with open("./data/cox-violent-parsed.csv") as f:
    reader = PeekyReader(DictReader(f))
    try:
        while True:
            p = Person(reader)
            if p.valid:
                vpeople.append(p)
    except StopIteration:
        pass

vpop = list(filter(lambda i: ((i.violent_recidivist == True and i.lifetime <= 730) or
                              i.lifetime > 730), list(filter(lambda x: x.vscore_valid, vpeople))))

vrecid = list(filter(lambda i: i.violent_recidivist == True and i.lifetime <= 730, vpeople))

vrset = set(vrecid)

vsurv = [i for i in vpop if i not in vrset]
```

```{python}

create_table(vrecid, vsurv).to_csv("data/table_vrecid.csv")

```

```{r}

read.csv(here("data", "table_vrecid.csv"))[,-1] %>%
  ggplot(aes(x = Metrics, y = Scores)) +
  geom_col() +
  labs(title = "Violent recidivism")

```


Even more so for Black defendants.

````{python}

create_comp_tables(vrecid, vsurv).to_csv("data/comp_tables_vrecid.csv")

```

```{r}

read.csv(here("data", "comp_tables_vrecid.csv"))[,-1] %>%
  ggplot(aes(x = Metrics, y = Scores, fill = Group)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(title = "Violent recidivism")

```












